---
title: DS202 Lab 5
author: Aaron Jencks, Eli Musgrove
date: 4/11/2020
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(readr)
library(plotly)
library(lubridate)
```

## Data Background

```{r}
acc = read_csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv")  #, stringsAsFactors = FALSE)
person = read_csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv")  #, stringAsFactors = FALSE)

str(acc)
str(person)
```

## Part One

1. Are there some days of the week where more accidents happen than on others?

```{r}
ggplot(acc, aes(x=DAY_WEEK)) + geom_bar()
```

There are definitely more accidents on the weekends than on other days of the weeks, thought, it would seem that Thursday and Friday also represent slightly higher counts than others.

2. What time of the day to accidents happen?

```{r}
acc = acc %>% filter(HOUR < 25)
ggplot(acc, aes(x=HOUR)) + geom_bar()
```

There seems to be more accidents the later in the day it gets, right up to about 6p, then it starts to decline. There are also spikes in the earlier hours when people leave to go to work, and when they break for lunch.

3. What is the number of accidents with at least one drunk driver?

```{r}
ggplot(acc, aes(x=DRUNK_DR)) + geom_bar()
sum(acc$DRUNK_DR > 0)
```

There are roughly 8667 accidents involving at least one drunk driver

## Part Two

4. In the `person` table, identify drivers and subset them

```{r}
# names(person)
# names(acc)
drivers = person %>% filter(PER_TYP == 1)
# str(drivers)
```

5. Join accident and person table.

```{r}
joined = left_join(acc, person, by = "ST_CASE")
str(joined)
```

6. Tally the number of accidents by day of the week and gender. Visualize the results.

```{r}
summ = joined %>% mutate(SEX = factor(SEX, levels=c(1, 2), labels=c('MALE', 'FEMALE'))) %>% filter(!is.na(SEX)) %>% group_by(DAY_WEEK, SEX) %>% summarize(count=n())
ggplot(summ, aes(x=DAY_WEEK, y=count, fill=SEX, label=count)) + geom_bar(stat="identity", position="dodge") + geom_text(vjust=-0.25)
```

On average, there seems to be at least twice as many accidents caused my men, than there are women, like almost exactly twice as many, a little less on Thursday.

## Part Three

7. Now plot a chloropleth map of accidents on a county level
	- Read the description for `STATE` and `COUNTY` columns in the manual. The state & county codes are Geographic Locator Codes from [the General Services Acministration's GSA publication](https://www.gsa.gov/reference/geographic-locator-codes/glcs-for-the-us-and-us-territories)
	- Use `readxl::read_xlsx` to read in the GLCs.
8. Visualize the number of accidents in each county using a chloropleth map. To do this, you need to join the accident data, GLCs, and appropriate map date.
9. Can you find seasonal trends in accident occurrences in different states? Is summer or winter more dangerous? Why?
